SimBio: a versatile and expandable Python package for modeling systems.

### 2023-11-14 article.qmd
Add Hernan's changes

---
title: "SimBio: a flexible and extensible Python package for systems modelling"
bibliography: bibliography.bib
format:
  html:
    number-sections: true
    anchor-sections: true
  pdf:
    documentclass: scrartcl
    papersize: a4
    geometry:
        - margin=20mm
    pdf-engine: tectonic
    number-sections: true
    colorlinks: true
---

@@

Chemical Reaction Networks (CRN)
are the basis of many models used across the biological sciences and epidemiology.
They consist of a set of species and reactions that mediate conversions between them.
Several modalities exist to simulate these systems:
deterministic Ordinary Differential Equations (ODE) to model a macroscopic behavior,
Stochastic Differential Equations (SDE) to model microscopic fluctuations,
and jump processes (Gillespie-like simulations) to take into account the discreteness of the populations.
Instead of writing the equations directly for each of these formulations
which is error-prone and hard to reuse,
these models can be written in a higher level description
which can be compiled for these different types of simulations.

@@

A limitation of these tools is their extensibility from Python,
as they are or use wrapped libraries in other languages.
As such, they allow to "control" the model definition and simulation in a Python script,
but we cannot leverage other packages from the huge Python community
to change the way the simulation is carried out.
In some sense, we could say that
they facilitate not having to export the simulation result to a CSV file
and then loading it in Python to continue the analysis.

Another weak point is related to the way the models are written.
While most tools support standards such as the System Biology Markup Language (SBML) [@huckaSBMLL3V2],
no one is expected to write an SBML file directly.
Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
which can be translated to SBML.
A DSL permits to reuse of the code *as is* in different programming environments,
but will not be recognized in Integrated Development Enviromnents (IDE)
and, then, cannot provide the development experience we have come to expect:
syntax highlighting, code completion, refactoring, static analysis, among others.
This has been recognized by the community,
and an extension for Visual Studio Code has been created as part of a master thesis [@vscode-antimony].
But maintaining it into the future could be burdensome for the smaller systems biology community.

To address these two issues,
we developed SimBio,
a Python package to define and simulate reaction networks.
From the end-user point of view,
it is focused on providing an ergonomic experience
by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
From the power-user point of view,
since its coded in pure Python,
it can be extended from Python.
Its models can be introspected to create other representations
such as graphs connecting species and/or reactions,
or tables with parameters or equations.
Finally,
from the developer/maintainer point of view,
it has a modular architecture with a clear separation of concerns into three packages,
making it easier to maintain or to contribute new code.

@@

Modular code architecture enhances reusability, extendability and maintainability for the community.
Hence, we split the code to define and simulate reaction systems into three Python packages:
symbolite, to create symbolic expressions;
poincaré, to define dynamical systems;
and simbio, to define reaction systems
and interface with systems biology standards such as SBML.

@@

In contrast to previous tools,
both model definition and compilation into an ODE function is implemented in Python.
This lowers the barrier to develop different simulation modalities,
such as perfomance optimizations by leveraging some structure in the model.
But being in Python does not mean less performance than C/C++ tools,
as the resulting ODE function can be *jitted* using numba.

These set of packages have a modular architecture with a clear separation of concerns.
As such, they can be reused, improved and extended by the broader Python community.
For example, people outside the systems biology community could
add a stochastic integrator to poincaré,
which could then be used from SimBio.
This architecture also makes them easier to understand,
reducing the difficulty to contribute fixes or new code,
assuring their maintainance and continued development into the future.

>>>>

---
title: "SimBio: a versatile and expandable Python package for modeling systems."
bibliography: bibliography.bib
format:
  html:
    number-sections: true
    anchor-sections: true
  pdf:
    documentclass: scrartcl
    papersize: a4
    geometry:
        - margin=20mm
    pdf-engine: tectonic
    number-sections: true
    colorlinks: true
  docx:
    number-sections: true
---

@@

Chemical Reaction Networks (CRN)
are the foundation of many models used in biological sciences and epidemiology.
They are made up of a set of species and reactions that mediate conversions between them.
There are several ways to simulate these systems:
deterministic Ordinary Differential Equations (ODEs) to model macroscopic behavior,
Stochastic Differential Equations (SDEs) to model microscopic fluctuations,
and jump processes (Gillespie-like simulations) to account for the discreteness of populations.
Instead of writing the equations directly for each of these formulations
which is error-prone and difficult to reuse,
these models can be written in a higher-level description
that can be compiled for these different types of simulations.

@@

One limitation of these tools is that they are not extensible from Python,
as they are either built on or use wrapped libraries in other languages.
This means that they allow you to "control" the model definition and simulation in a Python script,
but you cannot use other packages from the vast Python community to change the way the simulation is carried out.
In a way, we could say that
they make it easier to avoid having to export the simulation results to a CSV file
and then load them into Python for further analysis.
TODO: GIVE EXAMPLES

Another weakness is related to the way models are written.
While most tools support standards such as the System Biology Markup Language (SBML) [@huckaSBMLL3V2],
no one is expected to write an SBML file directly.
Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
which can be translated to SBML.
A DSL allows code to be reused in different programming environments,
but it will not be recognized by Integrated Development Environments (IDEs)
and therefore cannot provide the development experience we have come to expect, such as
syntax highlighting, code completion, refactoring, and static analysis.
This has been recognized by the community,
and an extension for Visual Studio Code has been created as part of a master thesis [@vscode-antimony].
However, maintaining it in the future could be burdensome for the smaller systems biology community.

We developed SimBio,
a Python package for defining and simulating reaction networks,
to address these two issues.
It focuses on providing an ergonomic experience for end-users
by integrating well with IDEs and static analysis tools
through the use of standard modern Python syntax.
It can be extended from Python
since it is coded in pure Python,
which is beneficial for power users.
Its models can be introspected to create other representations,
such as graphs connecting species and/or reactions,
or tables with parameters or equations.
Finally,
it has a modular architecture with a clear separation of concerns into three packages,
making it easier to maintain or to contribute new code,
which is beneficial for developers and maintainers.

@@

Modular code architecture makes code reusable, extensible, and easier to maintain.
Therefore, we split the code to define and simulate reaction systems into three Python packages:
symbolite, to create symbolic expressions;
poincaré, to define dynamical systems;
and simbio, to define reaction systems and interface with systems biology standards such as SBML.
TODO: NOT SURE SUBSECTIONS ARE A GOOD IDEA. MAYBE JUST BOLD

@@

Unlike previous tools,
both model definition and compilation into an ODE function are implemented in Python.
This lowers the barrier to developing different simulation modalities,
such as performance optimizations by leveraging some structure in the model.
However, being in Python does not mean that performance is less than that of C/C++ tools,
as the resulting ODE function can be JITted using Numba.

These packages are designed with a modular architecture that clearly separates different concerns.
This makes them easier to reuse, improve, and extend by the broader Python community.
For example, someone outside the systems biology community could add a stochastic integrator to poincaré,
which could then be used from SimBio.
This architecture also makes them easier to understand,
which reduces the difficulty of contributing fixes or new code,
and ensures their maintenance and continued development into the future.
